\section{\F: Command Shell}\label{pyforth}

Program works, but we need something more interesting. Lets do some
\emph{interactive system}, in parralel \emph{adding some control constructions}
to our assembler. At this point we
\begin{itemize}
	\item have no tools let as define complex syntax in our language
	\item we dont want to mess two parsers: assembler and command shell
\end{itemize}
So we need something very strange for our command shell: programming language
without syntax. In fact it is impossible, but there is one language with ultra
simple syntax which \emph{parser can be rewritten in few machine commands}: \F.

If you know something about \F, it is not suprize for you that we use it: our
\term{virtual machine}
was designed very close to this language principles. We do some shift
from original \F, as we want to manipulate with objects, but not with bytes
and machine integers (see next page), as you do making original
\term{Virtual FORTH Machine}.

\bigskip
Returning to \F\ syntax, it is very simple:
\begin{enumerate}[nosep]
\item collect sequential symbols one by one from input stream until first
space encountered\ --- \emph{tadaam! thats all: you have got the parser 
finished 8-)}
\item and try to do something with selected string (not surprizing that this 
string has special name in \F\ --- \term{word}).
\end{enumerate}
On next stage \F-system tries to find this word in special structure\ ---
\term{vocabulary}. Vocabulary is linked list contains compiled procedures
with its names stored in.
If parsed word matches procedure name, this procedure will
be executed immediately. If word not found in vocabulary, \F\ tries to convert
it into integer number\note{there is no floating point support in core
\F\ language\ --- are you already scared?}, and push into \term{data stack}.

\bigskip
We are not going to write yet another tutorial on \F\ here. It is better
if you take a break here, and look into first chapters of real \emph{cool book:
Leo Brodie's Starting \F}\ \cite{starting}. You will be amazed by \F\ miracle
simplicity, and scared by its dragon ass \cite{dragon}\ of zen syntax,
data stack bitbanging, and functional limitations\note{no float numbers,
no files, do you want data structures? make arrays yourself!}.

\bigskip
Also we will do some ANS Forth standard sidestep, see notes.

\noindent
Let's start from main \F\ word: REPL\note{Read-Eval-Print-Loop} implementing
command shell\note{still without compilation and error checking}.
\begin{lstlisting}[language=Forth]
: INTERPRET		\ REPL interpreter loop
	begin
		\ get next word name from input stream
		word	( -- str:wordname ) 
		\ find word entry point
		find	( str:wordname -- addr:xt )
		\ execute found xt (execution token) 
		execute	( xt -- ) 
	again
;
\end{lstlisting}
Don't forget how we should pass this source for compilation and run:
enclose it into \verb|VM(''' ... code ... ''')| in \py\ source.

\subsection{Expanding VM using inheritance}

A current parser will not process this source, as syntax notably changed.
We will not break VM demo parser, but use \emph{VM class inheritance to redefine
parser} for special \F\ dialect, and add some extra command to original VM.

When you are going to implement your own language on top of VM, you should 
go this way too:
\begin{itemize}
\item fork special class from core VM
\item implement parser/compiler for your language
\item \emph{expand inherited VM} with specific commands your need of
\end{itemize}
Last item must be focused: \emph{don't change core VM class} with new commands
or modified behaviour, as this changes will impact on all other inherited VMs.
We broke this rule here, only because VM can't do anything in this point, and we
need a lot of to add to make core VM be usable itself.

\begin{lstlisting}[language=Python]
# it should look like this:
class FORTH(VM):
	t_ignore_COMMENT = r'\#.*|\\.*|\(.*?\)'		# comment
\end{lstlisting}

And here we have a problem, Houston! All elements of parser are encapsulated
in \verb|compiler()| method using closure, and oops
we \emph{must rework VM class} itself.
The way we can do it you can see on this short sample. This is
lexer-only variant of simplest Hello World parser uses inheritance:
\lstinputlisting[language=python]{../inher.py}

I faced with problem in PLY: its lexer generator have problem with
\verb|t_error| unicality validation, if there are more \verb|t_error| functions
(or methods) defined anywhere in one module. So I made tiny fix in
\verb|ply/lex.py|, see
\href{http://github.com/dabeaz/ply/issues/142}{issue 142}, but my pull request
still not merged, and \emph{you must clone patched PLY right here}:
\begin{lstlisting}
~/DLR$ git clone -o ponyatov \
	https://github.com/ponyatov/ply.git
\end{lstlisting}
For import patched PLY we need to push it in search path \emph{before}
\verb|/usr/lib|:
\begin{lstlisting}[language=Python]
import sys ; sys.path.insert(0,'./ply')
import ply.lex  as lex ; ply_class_inherit = True # use fix
\end{lstlisting}

\bigskip
Going to rewrite VM, we need \verb|t_rules| as class static strings, and
methods, moved out of \verb|compiler()|:
\begin{lstlisting}[language=Python]
class VM:
	# ===== lexer code section =====
	t_ignore = ' \t\r'				# drop spaces (no EOL)
	t_ignore_COMMENT = r'\#.*'		# line comment
	# regexp/action rules (ANY state)
	def t_ANY_newline(self,t):		# special rule for EOL
		r'\n'
		t.lexer.lineno += 1			# increment line counter
		# do not return token, it will be ignored by parser
	# token types
	tokens = ['NOP','BYE','REGISTER','EQ','STRING']
	# required lexer error callback must be method
	def t_ANY_error(self,t): raise SyntaxError('lexer: %s' % t)
\end{lstlisting}
Expand comment regexp and tokens set in FORTH:
\begin{lstlisting}[language=Python]
class FORTH(VM):
	t_ignore_COMMENT = r'\#.*|\\.*|\(.*?\)'		# comment
	tokens = ['NOP','BYE','REGISTER','EQ','STRING',
			'COLON','SEMICOLON','BEGIN','AGAIN']	# : ;
\end{lstlisting}
Change lexer creation:
\begin{lstlisting}[language=Python]
class VM:
	def compiler(self,src):
		...
		# create lexer object
		# note: we point on object (instance!) with rules 
		self.lexer = lex.lex(object=self)
		# feed source code
		self.lexer.input(src)
\end{lstlisting}
and parser the same:
\begin{lstlisting}[language=Python]
class VM:
	def compiler(self,src):
		# create ply.yacc object, without extra files
		parser = yacc.yacc(debug=False,write_tables=None,\
			module=self) # here we must point on instance
		# parse source code using lexer
		parser.parse(src,self.lexer)
\end{lstlisting}

In \F\ syntax we have no ability to distinct between VM commands, defined in
vocabulary and yet undefined words using only regexp. So we need to group tokens
in one ID universal token, and do some lookup to detect what token type we have
in every point.

\begin{lstlisting}[language=Python]
class FORTH(VM):
	tokens = ['ID','COLON','SEMICOLON','BEGIN','AGAIN']
\end{lstlisting}
We removed some token types, and PLY will give as report\note{PLY was developed
by David M. Beazley specially for student learning purposes, so good verbosity
is key feature of this parser generator. As we are going to build power
\emph{metaprogramming system}, which can be thinked as \textit{interactive
compiler \& algorithm development environment}, we should go this way too
later making parser generators and other tools, visualization and verbosity must
have.}\ what's wrong:
\begin{lstlisting}
ERROR: /home/ponyatov/DLR/VM.py:62:
	Symbol 'NOP' used, but not defined as a token or a rule
ERROR: /home/ponyatov/DLR/VM.py:65:
	Symbol 'BYE' used, but not defined as a token or a rule
ERROR: /home/ponyatov/DLR/VM.py:68:
	Symbol 'REGISTER' used, but not defined as a token or a rule
ERROR: /home/ponyatov/DLR/VM.py:68:
	Symbol 'EQ' used, but not defined as a token or a rule
ERROR: /home/ponyatov/DLR/VM.py:73:
	Symbol 'STRING' used, but not defined as a token or a rule
WARNING: Token 'AGAIN' defined, but not used
WARNING: Token 'BEGIN' defined, but not used
WARNING: Token 'SEMICOLON' defined, but not used
WARNING: Token 'COLON' defined, but not used
WARNING: Token 'ID' defined, but not used
WARNING: There are 5 unused tokens
ERROR: Infinite recursion detected for symbol 'constant'
ERROR: Infinite recursion detected for symbol 'command'
ply.yacc.YaccError: Unable to build parser
\end{lstlisting}
Here we see source line numbers, points directly on problematic gramma rules,
and list of unused tokens.

So we need to drop corresponding rules
from inherited class.
The second tiny patch i've done for PLY, lets mark parser rules as deleted using
\verb|=Null| assignment:
\begin{lstlisting}[language=Python]
class FORTH(VM):
	t_NOP = p_NOP = None
 	t_BYE = p_BYE = None
 	t_REGISTER = p_Rload
 	p_constant_STRING = None
\end{lstlisting}
\begin{lstlisting}
ERROR: /home/ponyatov/DLR/VM.py:60: Symbol 'command' used,
	but not defined as a token or a rule
\end{lstlisting}
We must define what commands we are going to see in our programs.
\begin{description}[nosep]
\item[grammar] rules modifies VM grammar, and required
\begin{lstlisting}[language=Python]
# grammar override
def p_ID(self,p):		' command : ID '
def p_COLON(self,p):	' command : COLON '
def p_SEMICOLON(self,p):' command : SEMICOLON '
def p_BEGIN(self,p):	' command : BEGIN '
def p_AGAIN(self,p):	' command : AGAIN '
\end{lstlisting}
\item[lexemes] we want to override comparing to base VM regexp rule set
\begin{lstlisting}[language=Python]
# lexemes regexp overrides
t_COLON = ':' ; t_SEMICOLON = ';'
def t_BEGIN(self,t):
	r'begin'
	return t 
def t_AGAIN(self,t):
	r'again'
	return t
def t_ID(self,t): # this rule must be last rule
	r'[a-zA-Z0-9_]+'
	return t 
\end{lstlisting}
\end{description}
AssertionError, we got empty program:
\verb|self.Ip < len(self.program)|

\subsection{Token overwrite for commands and \F\ words}

Let's add fix to begin of source lets run dummy program as usual:
\begin{lstlisting}[language=Python]
FORTH(r''' # use r' : we have escapes in string constants
nop bye
...
\end{lstlisting}
Base \verb|VM| compiler detects this commands directly in parser syntax rules,
and compiles them. In new inherited \verb|FORTH| machine we disabled this rules,
and moved all word-looking char sequences into one \verb|ID| token. In sample
source you can see that this ID can be new word name, VM command or control
structure statement\note{In \F\ all control structures must be implemented using
generic \F-words, we'll see how to make your own controls yourself later. But
now we will use \emph{parser-based method} to show you how to implement them in
your assembly or compiler}. In ID lex rule we can detect what is this ID in
real, and use \term{terminal overwrite}.

We have to add some hint let us check is given word name:
\begin{description}[nosep]
\item[VM command]\ \\
we can use \py\ \term{reflection} to look up is given string is FORTH
class method name or its VM superclass method, but for simplicity we will use
special dict:
\begin{lstlisting}[language=Python]
class FORTH(VM):
	# command lookup table: string -> method
	cmd = { 'nop':VM.nop , 'bye':VM.bye }
\end{lstlisting}
\item[already defined word] exists in vocabulary\\
it is intuitive we must have vocabulary itself:
\begin{lstlisting}[language=Python]
class FORTH(VM):
	# vocabulary of all defined words
	voc = {}
\end{lstlisting}
\item[undetected ID], maybe new word or some misspelling
\end{description}
Now we can add \term{lexeme overriding} into \verb|t_ID| lexer rule:
\begin{lstlisting}[language=Python]
	# add extra types for tokens
	tokens = ['ID','CMD','VOC',
				'COLON','SEMICOLON','BEGIN','AGAIN']
 	def t_ID(self,t): # this rule must be last rule
 		r'[a-zA-Z0-9_]+'
 		# first lookup in vocabulary
 		if t.value in self.voc: t.type='VOC'
 		# then check is it command name
 		if t.value in self.cmd: t.type='CMD'
 		return t 
\end{lstlisting}
Note checking order: this variant let as define new words with names equal to VM
commands, and later we'll do it in \term{\F-assembler} \ref{forthasm} let you
compile or execute single VM command in interactive \F\ system.

\clearpage
And define new special rules in grammar:
\begin{lstlisting}[language=Python]
 	def p_CMD(self,p):
 		' command : CMD '
 		# compile command using cmd{} lookup table
 		self.program.append(self.cmd[p[1]])
 	def p_VOC(self,p):		' command : VOC '
\end{lstlisting}

Dummy compiles and executes\note{check VM.interpreter() uses command(self)}, but
we need to \emph{run last defined word as program entry point}. To do this we
must implement all parser/compiler staff we did as stubs.

\clearpage
\subsection{Implement : COLON ; definition}

\begin{lstlisting}[language=Python]
class FORTH(VM):
	# vocabulary of all defined words
	voc = {}

  	def p_COLON(self,p):
  		' command : COLON ID'
  		# store current compilation pointer into voc
  		self.voc[p[2]] = len(self.program)
  		print self.voc
  	def p_ID(self,p):		' command : ID '
\end{lstlisting}
\begin{lstlisting}
{'INTERPRET': 2}
0000 <unbound method VM.nop> [0, 1, 2, 3, 4, 5, 6, 7]
0001 <unbound method VM.bye> [0, 1, 2, 3, 4, 5, 6, 7]
\end{lstlisting}
What we do? On \verb|: INTERPRET| code we \emph{marked current compiler
position} into vocabulary. Number 2 is address (index in \verb|program[2]|) of
command will be compiled just next.
\begin{lstlisting}[language=Python]
class FORTH(VM):
	# command lookup table
	cmd = { 'nop':VM.nop , 'bye':VM.bye , 'ret':VM.ret}
 	def p_SEMICOLON(self,p):
 		' command : SEMICOLON '
 		self.program.append(self.cmd['ret'])
 		
class VM:
  	def dump(self):	
  		print
  		for i in range(len(self.program)):
  			print '%.4X: %s' % (i,self.program[i])
  		print
	def __init__(self, P=''):
		self.compiler(P)			# run parser/compiler
		self.dump()					# dump compiled program
		self.interpreter()		  	# run interpreter
\end{lstlisting}
We added \verb|ret| command compilation runs on ; in source code. If you look on
program dump, you'll see that \verb|INTERPRET| item in vocabulary points on
address of first command, compiled by \verb|;| (out parser ignores all code
between :
; )
\begin{lstlisting}
{'INTERPRET': 2}
0000: <unbound method VM.nop>
0001: <unbound method VM.bye>
0002: <unbound method VM.ret>
\end{lstlisting}

\clearpage
\subsection{CALL and RET commands}
As you see from \cite{starting}\ and especially from \cite{threaded}, \F\
program is tangle of spaghetti \term{threaded code}\note{distinct this term from
\term{thread} term in \term{multitasking}}, where every word use other words
via nested \term{call}. Every time when call occurs, current execution
(interpreter) pointer pushes into special \term{return stack}, and jumps to
required word code. When called word finishes its work, \verb|ret| command pops
\term{return address} from stack, and \emph{return execution on next command}
after used \verb|call| command\note{there is some variant of pure threaded code
you can see in \cite{thbell}\ --- only addresses without call opcode, and
special address for ret}.
\begin{lstlisting}[language=Python]
class VM:

	D = []							# data stack
	R = []							# CALL/RET return stack
	def call(self):
		# push return address (Ip points to call parameter)
		self.R.append(self.Ip+1)
		self.Ip = self.program[self.Ip]	# jmp
	def ret(self):
		assert self.R					# check non-empty
		self.Ip = self.R.pop()	# return to marked address		

class FORTH(VM):
	# command lookup table
	cmd = { 'nop':VM.nop , 'bye':VM.bye ,
			'call':VM.call, 'ret':VM.ret}
 	def p_VOC(self,p):
		' command : VOC '
		self.program.append(VM.call);		# opcode
		self.program.append(self.voc[p[1]])	# cfa
\end{lstlisting}
Now we can implement and check calls for colon-defined words:
\begin{lstlisting}[language=Forth]
: NOOP ;
: INTERPRET		\ REPL interpreter loop
	NOOP
	...
\end{lstlisting}
\begin{lstlisting}
{'NOOP': 2, 'INTERPRET': 3}

	0000: <unbound method VM.nop> 
	0001: <unbound method VM.bye> 
NOOP: 
	0002: <unbound method VM.ret> 
INTERPRET: 
	0003: <unbound method VM.call> 0002 : NOOP 
	0005: <unbound method VM.ret>
\end{lstlisting}
To execute this we must add code, let us use last defined word as program
entry point. To do this, we replace first command with with \verb|call _entry|:

\begin{lstlisting}
	FORTH(r''' # use r' : we have escapes in string constants
\ test FORTH comment syntax for inherited parser
: NOOP ;
: INTERPRET		\ REPL interpreter loop
	NOOP
\end{lstlisting}
\begin{lstlisting}[language=Python]
class VM:
	def init_code(self): pass
		# set instruction pointer entry point
		self.Ip = 0							
		# clean up program memory	
		self.program = []
	def compiler(self,src):
		# ===== init code section =====
		self.init_code()

class FORTH(VM):
	def init_code(self):
		VM.init_code(self)
		self.program.append(VM.call)	# call ...
		self.program.append(0)			# _entry = 1
		self.program.append(VM.bye)		# bye
  	def p_COLON(self,p):
  		' command : COLON ID'
  		# store current compilation pointer into voc
		# reset _entry to current cfa
  		self.program[1] = self.voc[p[2]] = len(self.program)
		print self.voc
\end{lstlisting}

\subsection{Change tracing log output}

I have strage error in \verb|ret| command in \verb|NOOP|, so we need to change
tracing log generation: it must include both stacks, and have no info on
registers as \F\ don't use them.

Making program dump we need fast conversion from current address into record in
vocabulary. So \emph{we must use reversed vocabulary} to do reverse lookup from
addr to associated label (word name, variable,\ldots).
Also we need to change logging in interpreter: we don't need registers, but
need stacks to be printed on every command (if trace enabled).
\begin{lstlisting}[language=Python]
class VM:
	# return known label for given address
	def log_label(self,addr): return ''
	# dump command from given addr
	def log_command(self,addr):
		A = addr
		# current command
		command = self.program[A]
		# check if we have known labels
		L = self.log_label(A)
		if L: print '\n%s:' % L, # print on separate line
		# print main command log text
		print '\n\t%.4X' % A , command,
		# process commands with parameters
		if command == VM.call:
			# target address
			T = self.program[A+1]
			# print target addr with known label
			print T,self.log_label(T),
			A += 1
		return A+1	# return next command address
	# log extra state (in interpreter trace)
	def log_state(self):
		print self.Rg,
\end{lstlisting}
Note that \verb|log_command| \emph{returns address of next command}\ --- it is
required as \emph{we have commands with var length}. Interpreter don't use
returned addr, as it is done inside commands, but in \verb|dump| method every
next dumped addr must be computed in
\verb|log_command|.
\begin{lstlisting}[language=Python]
	def interpreter(self):
		self._bye = False		   				# stop flag
		while not self._bye:
			assert self.Ip < len(self.program)
			command = self.program[self.Ip]	# FETCH command
			self.log_command(self.Ip)		# log command
			self.log_state()				# log state
			self.Ip += 1				# to next command
			command(self)				# DECODE/EXECUTE
\end{lstlisting}
\begin{lstlisting}[language=Python]
   	def dump(self):
 		# loop over self.program
 		addr = 0;
 		# loop over program 
 		while addr < len(self.program):
 			# addr: command <extra>
 			addr = self.log_command(addr)
   		print ; print '-'*55
\end{lstlisting}
\begin{lstlisting}[language=Python]
class FORTH(VM):
	# vocabulary of all defined words
	voc = {}
	# reversed vocabulary {addr:name} for fast label lookup
	revoc = {}
	def log_state(self):
		print 'R:%s'%self.R,

  	def p_COLON(self,p):
  		' command : COLON ID'
  		# store current compilation pointer into voc
		# reset _entry to current cfa
  		self.program[1] = self.voc[p[2]] = len(self.program)
  		# add reversed pair {addr:label}
  		self.revoc[len(self.program)] = p[2]
		print self.voc
\end{lstlisting}
\begin{lstlisting}
{'NOOP': 3, 'INTERPRET': 4}
\end{lstlisting}
\clearpage
\begin{lstlisting}
	0000 <unbound method VM.call> 4 INTERPRET 
	0002 <unbound method VM.bye> 
NOOP: 
	0003 <unbound method VM.ret> 
INTERPRET: 
	0004 <unbound method VM.call> 3 NOOP 
	0006 <unbound method VM.ret>
-------------------------------------------------------
	0000 <unbound method VM.call> 4 INTERPRET R:[] 
INTERPRET: 
	0004 <unbound method VM.call> 3 NOOP R:[2] 
NOOP: 
	0003 <unbound method VM.ret> R:[2, 6] 
	0006 <unbound method VM.ret> R:[2] 
	0002 <unbound method VM.bye> R:[]
\end{lstlisting}
This trace log looks much better:
\begin{description}[nosep]
\item[first part]\ \\contains compiled program dump
\item[second part]\ \\contains interpreter execution log with stacks state
\emph{before command execution}
\end{description}

\bigskip
Also we can add extra check for call command: check that parameter after call
opcode was integer address:
\begin{lstlisting}[language=Python]
	def call(self):
		# push return address (Ip points to call parameter)
		self.R.append(self.Ip+1)
		self.Ip = self.program[self.Ip]			# jmp
		assert type(self.Ip) == int	# addr must be integer
\end{lstlisting}

\subsection{Throw out registers}

\F\ is pure stack language, and we will throw out registers\note{until we want
to play with self-made JIT}.
\begin{lstlisting}[language=Python]
class FORTH(VM):
	Rg = ld = None	# throw out registers
\end{lstlisting}

\subsection{Vocabulary structure}

You can see some strange \verb|cfa| label in listings above.
In \F\ there are some set of fields every word contains in vocabulary:
\begin{description}
	\item[LFA] \term{Link Field Area}\\
		In normal \F\ system words resides in memory in sequence.
		If we want to do search in vocabulary, we must have all words
		to be in \emph{linked list}, from end of vocabulary down to first word.
		So every LFA holds address of previous word.
		It is simplest way to do list in low level\ --- only single pointer
		(memory address) on previous element, followed by any data.

	\item[NFA] \term{Name Field Area}\\
		NFA holds word name. It can be any string type you use in all system.

	\item[AFA] \term{Attribute Field Area}\\
		You can add some attributes and flags to every word,\\
		\verb|IMMED| flag is only required (execute word in compilation mode).

	\item[CFA] \term{Code Field Area}\\
		is word code section,\\
		in other words it is \emph{address of first executable command}
		of code section

	\item[PFA] \term{Parameter Field Area} (optional)\\
		Some elements like constants and variables must hold some data,
		and PFA is memory area for this case. In CFA code section there is
		some code do something with PFA, like fetching data from PFA,
		or just pushing PFA address on data stack.
		
\end{description}

\noindent
Our \F-like system not \F\ in general, and we store vocabulary in
special data structure \emph{out of program memory}. Using \py\ dict
for vocabulary gives more speed search using hashed keys, but have some limits:
\begin{itemize}[nosep]
	\item dict can have only one key, and we use word names as keys, so we
		can't hold multiple (overrided) words with equal names in vocubulary
	\item data in dict not ordered, and we can't control search order, but it
		is not a matter, as we can't use overrided words in any way
\item \emph{vocabulary not in program address space}, and we can't hack 
		vocabulary itself using just memory manipulation. If we'll need this
		later, we must add some special commands to manipulate vocabulary.
\end{itemize}

\subsection{\F\ system stability}

Vocabulary separated out of main program memory is our first step to make
system much more stable then generic \F. In normal \F-system any accident write
to addressable memory\note{it is most frequent error linked with \emph{improper
operand order} on data stack, as \emph{bad address} computed by previous
code}\ can drop whole system in unstable crashing state and can broke your data.

Separating address spaces isolates this crashing, and system equipped with
debugger or fault protection subsystem, can do recovery or some emergency
processing before everything drops to hell.
Later we'll introduce OOP implementation, and \term{address isolation} between
objects will be good for stability.

\subsection{JMP command and BEGIN/AGAIN infinity loop}

In \F\ we want REPL\note{Read Eval Print Loop} loop as user command interface:
system must
\begin{enumerate}[nosep]
\item get source word by word separated by spaces,
\item do search in vocabulary, and 
\item execute every found word 
\item or print error message.
\item and repeat it infinity
\end{enumerate}
\noindent
To do infinity loop we need to implement begin/again pair in compiler.
At low level:
\begin{description}[nosep]
\item[begin] \emph{mark}s current compiler position in some
		\term{control compiling stack},\\
		this \emph{stack is required to let nested control structures}
	\item[again] \emph{resolve} saved position popping addres from stack, and 
		compiles \verb|JMP| command equal to \verb|goto| in high-level languages
\end{description}

We can use special dedicated control compiler stack, but we will use
\term{return stack}, as it is not used in compiling stage before interpreter
started. The
other reason to do this is to show how control structures are implemented in 
ordinary \F\ system\ --- they uses \term{return stack} for mark/resolve addresses
in controls compilation.

\begin{lstlisting}[language=Python]
class VM:

	def jmp(self):
		# get addr from jmp parameter
		self.Ip = self.program[self.Ip]
		# check type: addr must be integer
		assert type(self.Ip) == int
		# check range
		assert self.Ip < len(self.program)	

	def log_command(self,addr):
		if command in [ VM.jmp, VM.call]:
			print T,self.log_label(T),
class FORTH(VM):
	cmd = { 'jmp':VM.jmp, 'call':VM.call, 'ret':VM.ret,..}
\end{lstlisting}

\begin{lstlisting}[language=Python]
 	def p_BEGIN(self,p):
		' command : BEGIN '
		# mark Ip pushing in return stack
		self.R.append(len(self.program))
 	def p_AGAIN(self,p):
		' command : AGAIN '
		# jmp opcode
		self.program.append(self.cmd['jmp'])
		# jmp parameter: pop marked Ip
		self.program.append(self.R.pop())
\end{lstlisting}

\begin{lstlisting}
INTERPRET:
        0006 <unbound method VM.call> 5 NOOP
        0008 <function word at 0x018A51B0>
        0009 <function find at 0x018A51F0>
        000A <unbound method VM.jmp> 8
        000C <unbound method VM.ret>
\end{lstlisting}

\noindent
(*) \verb|word| and \verb|find| will be defined later.
\bigskip

As you see in listing, there is one \verb|jmp| command compiled by \verb|again|
at address \verb|0x000A|, jumps to \verb|0x0008| marked by \verb|begin| between
\verb|NOOP| and word in source code.

\clearpage
\subsection{Parsing input stream and search in vocabulary}

\begin{lstlisting}[language=Python]
class FORTH(VM):
	PAD = list('hello world ')
	def word(self):
		# result string
		S = ''
		# spaces
		BL = ' \t\r\n'
		# skip leading spaces
		while True:
			# pop first char from input stream
			C = self.PAD.pop(0)
			# break loop on non-space char
			if C not in BL: break
		# save first found non-BL char
		S += C
		# collect until BL char
		while True:
			C = self.PAD.pop(0)
			if C in BL: break
			S += C
		# push collected string
		self.D.append(S)
	
	def find(self):
		# pop word name from data stack
		S = self.D.pop()
		try:
			# push cfa of found word
			self.D.append(self.voc[S])
			# push FOUND flag
			self.D.append(True)
		except KeyError:
			# on error push word name
			self.D.append(S)
			# push NOT FOUND
			self.D.append(False)

	cmd = { 'word':word, 'find':find , ...}
\end{lstlisting}
You can fix parser: raise error on undefined commands/words:
\begin{lstlisting}[language=Python]
  	def p_ID(self,p):
		' command : ID '
#		raise BaseException(p[1])
\end{lstlisting}

If we change source:
\begin{lstlisting}[language=Forth]
: hello ;
: NOOP ;
: INTERPRET		\ REPL interpreter loop
	NOOP
	begin word find again
;
\end{lstlisting}
we'll get trace with found \verb|hello| and undefined \verb|world|:
\begin{lstlisting}
0007 <word at 0x018F51B0> D:[] R:[2]
0008 <find at 0x018F51F0> D:['hello'] R:[2]
0009 <method VM.jmp> 7  D:[3, True] R:[2]
0007 <word at 0x018F51B0> D:[3, True] R:[2]
0008 <find at 0x018F51F0> D:[3, True, 'world'] R:[2]
0009 <method VM.jmp> 7  D:[3, True, 'world', False] R:[2]
0007 <word at 0x018F51B0> D:[3, True, 'world', False] R:[2]
\end{lstlisting}

\subsection{IF/ELSE/ENDIF and EXECUTE/ABORT}

\begin{lstlisting}[language=Forth]
: INTERPRET				\ REPL interpreter loop
	begin
		\ get next word name from input stream
		word	( -- str:wordname )
		\ find word entry point
		find 	( addr:cfa true | str:wordname false )
		if ( addr:cfa )
			\ call to addr from stack
			execute	( addr:cfa -- )
		else ( str:wordname )
			\ dump state, stacks and restart
			abort
		endif		
	again			;		: hello ; \ defined word
\end{lstlisting}

\clearpage
\verb|execute| command works as \verb|call| but uses address from data stack:

\begin{lstlisting}[language=Python]
class VM:
	def execute(self):
		# push ret addr
		self.R.append(self.Ip)					
		# load jmp addr from stack
		assert self.D ; self.Ip = self.D.pop()	
		# addr must be integer
		assert type(self.Ip) == int				
		# check range
		assert self.Ip < len(self.program)		
class FORTH(VM):
	cmd = { 'execute':VM.execute,..} 
\end{lstlisting}

\clearpage
\verb|abort| must restart REPL loop, but we will stop all system with state
dump:
\begin{lstlisting}[language=Python]
class VM:
	def abort(self):
		print '\n\nD:%s\nR:%s\n' % (self.D,self.R)
		self.dump()
		raise BaseException('ABORT')
class FORTH(VM):
	cmd = { 'abort':VM.abort,..} 
\end{lstlisting}
\begin{lstlisting}
D:[3, 'world']
R:[2]

{'INTERPRET': 3}
	0000 <unbound method VM.call> 3 INTERPRET 
	0002 <unbound method VM.bye> 
\end{lstlisting}
		
\clearpage
\noindent
\verb|if|/\verb|else| is complex to understand: it uses forward jumps and
\term{backpatching} with \emph{one pass compiling}:
\begin{description}
\item[?jmp] conditional \verb|?jmp| command uses flag from data stack and
\emph{jumps on false}
\item[if] compiles \verb|?jmp| with \emph{fake} parameter, and \term{marks} its
parameter addr into control\note{don't forget we use return stack, but it can be
separate control stack}\ stack. Later this \verb|addr| will be popped by
\verb|else|/\verb|endif| and \term{backpatched} with real address must be
jumped if \emph{false} on data stack.
\item[endif] patches \verb|?jmp| command compiled by \verb|if| using
\verb|addr| popped from control stack
\item[else] (optional) implements else branch, compiles \verb|jmp endif|
command, replacing \verb|if addr| on constrol stack by \verb|jmp else| addr.
\end{description}

\begin{lstlisting}[language=Python]
class VM:
	def qjmp(self):
		assert self.D
		if not self.D.pop(): # ( bool:FALSE! -- ) 
			self.jmp()
		else:
			self.Ip += 1	# skip ?jmp parameter
			
	def log_command(self,addr):
		if command in [ VM.jmp, VM.qjmp, VM.call]:
			print '%.4X'%T,self.log_label(T),

class FORTH(VM):
	cmd = { 'jmp':VM.jmp, '?jmp':VM.qjmp,..}
\end{lstlisting}

\clearpage
\begin{lstlisting}[language=Python]
class FORTH(VM):

	# lexer
	tokens = [..,'IF','ELSE','ENDIF']
 	def t_IF(self,t):
 		r'if'
 		return t
 	def t_ELSE(self,t):
 		r'else'
 		return t
 	def t_ENDIF(self,t):
 		r'endif'
 		return t

 	# parser
	def p_IF(self,p):
		' command : IF '
		self.program.append(self.cmd['?jmp'])# opcode
		self.R.append(len(self.program))	# mark
		self.program.append(-1)				# fake addr
	def p_ELSE(self,p):
		' command : ELSE '
		assert self.R ; A = self.R.pop()	# pop if jmp
		self.program.append(self.cmd['jmp'])# jmp endif
		self.R.append(len(self.program))	# mark
		self.program.append(-1)				# fake addr
		self.program[A] = len(self.program)	# backpatch if
	def p_ENDIF(self,p):
		' command : ENDIF '
		assert self.R ; A = self.R.pop()	# resolve
		self.program[A] = len(self.program)	# backpatch
# 		self.program.append(self.cmd['nop'])# show endif pos
\end{lstlisting}

\begin{lstlisting}
INTERPRET: 
	0004 <function word at 0x7f31eece71b8> 
	0005 <function find at 0x7f31eece7230> 
	0006 <unbound method VM.qjmp> 000B  			# if
	0008 <unbound method VM.execute> 
	0009 <unbound method VM.jmp> 000C				# else
	000B <unbound method VM.abort> 
	000C <unbound method VM.nop>					# endif
	000D <unbound method VM.jmp> 0004 INTERPRET 
	000F <unbound method VM.ret>
\end{lstlisting}
Special \verb|nop| compilation was added to \verb|p_ENDIF| to show position
where \verb|endif| takes an action. In real code this position will used by 
first command compiled after \verb|endif|.

\clearpage
\begin{lstlisting}[language=Forth]
: hello if bye endif ;

hello: 
	0003 <unbound method VM.qjmp> 0006		\ if
	0005 <unbound method VM.bye> 
	0006 <unbound method VM.ret>			\ endif
\end{lstlisting}
This sample shows if/endif construct without else. And more complex nested
sample \verb|: hello if(1) nop else(1) if(2) bye endif(2) endif(1) ;|

\begin{lstlisting}[language=Forth]
	0003 <unbound method VM.qjmp> 0008		\ if(1)  
	0005 <unbound method VM.nop> 
	0006 <unbound method VM.jmp> 000B  		\ else(1)
	0008 <unbound method VM.qjmp> 000B		\ if(2)
	000A <unbound method VM.bye> 			\ endif(2)
	000B <unbound method VM.ret>			\ endif(1)
\end{lstlisting}
	